---
title: "11: Generalized Linear Models"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Describe the components of the generalized linear model (GLM)
2. Apply special cases of the GLM to real datasets
3. Interpret and report the results of GLMs in publication-style formats

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

# Set date to date format
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, 
                                               format = "%Y-%m-%d")

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## SIMPLE AND MULTIPLE LINEAR REGRESSION
The linear regression, like the t-test and ANOVA, is a special case of the **generalized linear model** (GLM). A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero. OR
2. The slope (beta) is equal to zero (typically used more)

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

### Simple Linear Regression
For the NTL-LTER dataset, can we predict irradiance (light level) from depth?
```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
# another way to format the lm function

irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
summary(irradiance.regression)
# y variables should always be on the left side, and x on the right
# Output tells us the irradiance at depth 0 is 487 (which = y intercept), and slope of line is -96 (so irradiance drops by 96 for each meter of depth)
# R squared gives you how well the trend can be explained by the linear regression
# Adjusted R-squared is 
# Multiple R-squared is 

# Correlation
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth)
# Tests correlation between 2 variables
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

> ANSWER: Irradiance is negatively correlated with depth (linear regression; R^2 = 0.31, df, = 15449, p < 0.0001)

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
plot(irradiance.regression)
# Not to be put in report, just used as an exploratory tool
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  ylim(0, 2000) +
  geom_point() 
print(irradiancebydepth) 

# Was it appropriate to use a linear regression here? No, because when we visualize our data, we see that it's actually following an exponential curve, not a linear one.
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values.

```{r, fig.height = 3, fig.width = 4}
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, irradianceWater != 0)
# We can't log transform 0s, so we have to remove them here (which is ok because its only 3 points, but its generally good practice to not delete data)

irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth)
summary(irradiance.regression2)
plot(irradiance.regression2)
# It's still a linear equation, but we can tell log transforming the data was better than not because our R squared is high (therefore the model does a good job of fitting the data)

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  # scale y log10 just puts our scale in the proper form
  geom_point() 
print(irradiancebydepth2) 

# SE can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point() +
    scale_y_log10() +
    geom_smooth(method = 'lm', se = FALSE, color = "blue")
print(irradiancebydepth2)

# Make the graph attractive
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point(alpha = 0.25) +
    scale_y_log10() +
    geom_smooth(method = 'lm', se = FALSE, color = "red") +
    labs(x = "Depth (m)", y = "Water Irradiance (units)") +
    scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12))
print(irradiancebydepth2)

```

### Non-parametric equivalent: Spearman's Rho
As with the t-test and ANOVA, there is a nonparametric variant to the linear regression. The **Spearman's rho** test has the advantage of not depending on the normal distribution, but this test is not as robust as the linear regression.

``` {r}
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth, 
         method = "spearman", exact = FALSE)
# Follow this code exactly to preform Spearman's Rho
```

### Multiple Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration could be dependent on depth and dissolved oxygen concentration: 

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = PeterPaul.chem.nutrients, tp_ug ~ depth + dissolvedOxygen)
summary(TPregression)
# Response on left, explanitory on right

TPplot <- ggplot(PeterPaul.chem.nutrients, 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)
# This shows us that the linear model is not a good predictor of phosphorus
```
### Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot")
library(corrplot)
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>%
  na.omit()
# It's important to omit NAs from entire dataset before running corrplot

PeterPaulCorr <- cor(PeterPaulnutrients)
# Find correlation between all elements of the dataset

corrplot(PeterPaulCorr, method = "ellipse")
# Ellipse helps you visualize the correlations better than the default (which is circles)

corrplot.mixed(PeterPaulCorr, upper = "ellipse")
```
### AIC to select variables

However, it is possible to over-parameterize a linear model. Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration.

```{r}
PeterPaul.naomit <- na.omit(PeterPaul.chem.nutrients)
TPAIC <- lm(data = PeterPaul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4)
# We're creating a linear regression to see how total phosphorus is related to all of the above. It's a lot of explanatory variables, too many for what we've been doing.

step(TPAIC)
# step generates AIC values (the lower the better), trying to find which explanatory variable(s) should be removed to get the lowest AIC value. Removing all but temp and total N gives us the lowest AIC, so that's what we'll run the regression with

TPmodel <- lm(data = PeterPaul.naomit, tp_ug ~ temperature_C + tn_ug)
summary(TPmodel)
# It's extremely important to understand your data so you end up with a model that makes sense
```

## ANCOVA
Analysis of Covariance consists of a prediction of a continuous response variable by both continuous and categorical explanatory variables. We set this up in R with the `lm` function, just like prior applications in this lesson. 

Let's say we wanted to predict total nitrogen concentrations by depth and by lake, similarly to what we did with a two-way ANOVA for depth ID and lake. 

```{r, fig.height = 3, fig.width = 4}
# main effects
TNancova.main <- lm(data = PeterPaul.chem.nutrients, tn_ug ~ lakename + depth)
summary(TNancova.main)
# Here we're asking whats the effect of lake name and whats the effect of depth on total N
# Output equation: TN = 353 + 135 - 9.7 + Error
# Output equation: y = (Paul @ 0) + (Peter @ 0) - (change in TN per unit change in depth) + Error

# interaction effects
TNancova.interaction <- lm(data = PeterPaul.chem.nutrients, tn_ug ~ lakename * depth)
summary(TNancova.interaction)
# TN = 325 + 185 - 20 - 48
# y = (Paul @ 0) + (Peter @ 0) - (change in TN in Paul lake per unit change in depth) + (change in TN in Peter lake per unit change in depth)

TNplot <- ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = tn_ug, color = lakename)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0, 10)
print(TNplot)
# ANCOVA makes different slopes for our different variables

# Make the graph attractive
```

